---
title: "Sampling Viral Shedding Trajectories"
subtitle: "Using the Bayesian joint model for agent-based simulation"
author: "Christopher Boyer"
date: today
format:
  html:
    embed-resources: true
    self-contained: true
    toc: true
    code-fold: false
    fig-width: 9
    fig-height: 5
    theme: cosmo
execute:
  warning: false
  message: false
---

## Overview

This document demonstrates how to use `sample_trajectories()` to generate
realistic SARS-CoV-2 viral shedding trajectories from the fitted Bayesian
joint model. Each trajectory provides daily values for:

| Output        | Description                                          |
|---------------|------------------------------------------------------|
| **RNA**       | Log viral RNA copies/mL (what PCR would measure)     |
| **PFU**       | Log PFU/mL (infectious virus, what culture measures)  |
| **LFD**       | Lateral flow device (rapid antigen test) result: 0/1 |
| **Symptomatic** | Whether the agent is symptomatic: 0/1             |

Trajectories span the **detectable period** (RNA or PFU above the limit of
detection) and reflect true inter-individual variability by sampling new
random effects from the fitted population distribution for each agent.

## Setup

```{r setup}
library(dplyr)
library(tidyr)
library(ggplot2)
library(cmdstanr)
library(targets)

# Load helper functions
source("R/utils.R")
source("R/data.R")
source("R/trajectories.R")

# Load fitted model and data
# Option A: if targets store works directly
# fit <- tar_read(kinetics_mcmc)

# Option B: reconstruct from CSV files (needed if paths have changed)
csv_files <- sort(list.files("output/stan_csv",
                             pattern = "4827d7[.]csv$",
                             full.names = TRUE))
fit <- as_cmdstan_fit(csv_files)

stan_data <- tar_read(stan_data)
```

## Example 1: Single agent, multiple draws

Generate 200 trajectories for a 35-year-old, unvaccinated person with a
first Omicron infection. Each trajectory represents one possible infection
episode — the spread across draws captures both posterior parameter
uncertainty and true biological inter-individual variability.

```{r single-agent}
agent <- data.frame(
  age_group       = "30-49",
  variant         = "omicron",
  vaccination     = "unvaccinated",
  prior_infection = FALSE
)

traj <- sample_trajectories(
  fit, stan_data, agent,
  n_draws = 200,
  seed    = 42
)

glimpse(traj)
```

### Viral load trajectories

```{r plot-single-vl}
#| fig-height: 5

# Pivot to long format, keeping only days where each marker is above its LOD.
# The detection window spans any day where RNA *or* PFU is detectable; PFU
# typically has a narrower window than RNA, so without filtering the PFU
# panel would show latent values deep below LOD.
traj_long <- traj |>
  select(draw, day_since_detectable, log_rna, log_pfu,
         rna_detectable, pfu_detectable) |>
  pivot_longer(c(log_rna, log_pfu),
               names_to = "marker", values_to = "value") |>
  mutate(
    detectable = if_else(marker == "log_rna", rna_detectable, pfu_detectable),
    marker = recode(marker,
      log_rna = "RNA (log copies/mL)",
      log_pfu = "PFU (log PFU/mL)"
    )
  ) |>
  filter(detectable)

ggplot(traj_long, aes(day_since_detectable, value, group = draw)) +
  geom_line(alpha = 0.08, colour = "steelblue") +
  facet_wrap(~marker) + #scales = "free") +
  labs(
    x = "Days since first detectable",
    y = "Viral load (log scale)",
    title = "200 sampled trajectories — single Omicron agent"
  ) +
  theme_minimal(base_size = 13)
```

### LFD positivity and symptom onset timing

```{r plot-single-outcomes}
#| fig-height: 4

outcome_summary <- traj |>
  group_by(draw) |>
  summarise(
    detectable_days   = n(),
    lfd_positive_days = sum(lfd),
    symptom_onset_day = if (any(symptomatic == 1))
      min(day_since_detectable[symptomatic == 1]) else NA_real_,
    ever_symptomatic  = any(symptomatic == 1),
    .groups = "drop"
  )

outcome_summary |>
  select(detectable_days, lfd_positive_days, symptom_onset_day) |>
  pivot_longer(everything(), names_to = "outcome", values_to = "value") |>
  mutate(outcome = recode(outcome,
    detectable_days   = "Detectable duration (days)",
    lfd_positive_days = "LFD-positive days",
    symptom_onset_day = "Symptom onset (day)"
  )) |>
  ggplot(aes(value)) +
  geom_histogram(bins = 20, fill = "steelblue", alpha = 0.7) +
  facet_wrap(~outcome, scales = "free") +
  labs(x = NULL, y = "Count",
       title = "Distribution of clinical outcomes across 200 trajectories") +
  theme_minimal(base_size = 13)
```

```{r}
cat(sprintf("Ever symptomatic: %d / %d (%.0f%%)\n",
            sum(outcome_summary$ever_symptomatic),
            nrow(outcome_summary),
            100 * mean(outcome_summary$ever_symptomatic)))
```


## Example 2: Comparing agent profiles

How do host characteristics shape viral kinetics? Let's compare three
distinct profiles:

```{r multi-agent}
agents <- data.frame(
  label           = c("Young, Omicron, unvaxxed",
                       "Middle-aged, Delta, boosted",
                       "Older, Omicron, boosted + reinfection"),
  age_group       = c("0-29",  "30-49", "50+"),
  variant         = c("omicron", "delta", "omicron"),
  vaccination     = c("unvaccinated", "boosted", "boosted"),
  prior_infection = c(FALSE, FALSE, TRUE)
)

traj2 <- sample_trajectories(
  fit, stan_data, agents,
  n_draws = 200,
  seed    = 123
)
```

### Trajectory envelopes by profile

```{r plot-multi-envelope}
#| fig-height: 8

summary_rna <- traj2 |>
  filter(rna_detectable) |>
  group_by(label, day_since_detectable) |>
  summarise(
    rna_med = median(log_rna),
    rna_lo  = quantile(log_rna, 0.1),
    rna_hi  = quantile(log_rna, 0.9),
    n_draws = n(),
    .groups = "drop"
  ) |>
  filter(n_draws >= 0.2 * 200)

summary_pfu <- traj2 |>
  filter(pfu_detectable) |>
  group_by(label, day_since_detectable) |>
  summarise(
    pfu_med = median(log_pfu),
    pfu_lo  = quantile(log_pfu, 0.1),
    pfu_hi  = quantile(log_pfu, 0.9),
    n_draws = n(),
    .groups = "drop"
  ) |>
  filter(n_draws >= 0.2 * 200)

ggplot(mapping = aes(x = day_since_detectable)) +
  geom_ribbon(data = summary_rna, aes(ymin = rna_lo, ymax = rna_hi),
              alpha = 0.2, fill = "firebrick") +
  geom_line(data = summary_rna, aes(y = rna_med),
            colour = "firebrick", linewidth = 0.8) +
  geom_ribbon(data = summary_pfu, aes(ymin = pfu_lo, ymax = pfu_hi),
              alpha = 0.2, fill = "steelblue") +
  geom_line(data = summary_pfu, aes(y = pfu_med),
            colour = "steelblue", linewidth = 0.8) +
  facet_wrap(~label, ncol = 1) +
  labs(
    x = "Days since first detectable",
    y = "Viral load (log scale)",
    title = "Median and 80% interval: RNA (red) vs PFU (blue)"
  ) +
  theme_minimal(base_size = 13)
```

### Duration comparison

```{r plot-multi-duration}
#| fig-height: 4

traj2 |>
  group_by(label, draw) |>
  summarise(detectable_days = n(), .groups = "drop") |>
  ggplot(aes(x = detectable_days, fill = label)) +
  geom_histogram(bins = 20, alpha = 0.6, position = "identity") +
  labs(
    x = "Detectable duration (days)",
    y = "Count",
    fill = "Agent profile",
    title = "Detection window by agent characteristics"
  ) +
  theme_minimal(base_size = 13) +
  theme(legend.position = "bottom")
```


## Example 3: Latent trajectories (no measurement noise)

For transmission modelling, you may want the true underlying viral load
without measurement noise. Set `include_noise = FALSE` to get latent
trajectories and exact probabilities.

```{r latent}
traj_latent <- sample_trajectories(
  fit, stan_data, agents[1, , drop = FALSE],
  n_draws       = 30,
  include_noise = FALSE,
  seed          = 99
)
```

When `include_noise = FALSE`:

- `rna` and `pfu` equal the latent values (`log_rna`, `log_pfu`) — no
  Gaussian noise or LOD censoring
- `lfd` is the **probability** of a positive LFD (continuous, 0–1)
- `symptomatic` is the **daily hazard** of symptom onset (continuous, 0–1)

```{r plot-latent}
#| fig-height: 5

p1 <- ggplot(traj_latent, aes(day_since_detectable, lfd, group = draw)) +
  geom_line(alpha = 0.4, colour = "darkorange") +
  labs(x = "Days since detectable", y = "P(LFD positive)",
       title = "LFD positivity probability") +
  theme_minimal(base_size = 12)

p2 <- ggplot(traj_latent, aes(day_since_detectable, symptomatic, group = draw)) +
  geom_line(alpha = 0.4, colour = "purple") +
  labs(x = "Days since detectable", y = "Daily symptom hazard",
       title = "Symptom onset hazard") +
  theme_minimal(base_size = 12)

library(patchwork)
p1 + p2
```


## Example 4: Generating a population for ABM

To initialise 1,000 agents for an ABM, sample characteristics from a
distribution and call `sample_trajectories()` with `n_draws = 1` to get
one realised trajectory per agent.

```{r abm-population}
set.seed(2026)
n_pop <- 100  # use 1000+ for real ABM; 100 here for speed

pop <- data.frame(
  age_group = sample(c("0-29", "30-49", "50+"), n_pop,
                     replace = TRUE, prob = c(0.3, 0.4, 0.3)),
  variant   = "omicron",
  vaccination = sample(c("unvaccinated", "unboosted", "boosted"), n_pop,
                       replace = TRUE, prob = c(0.2, 0.3, 0.5)),
  prior_infection = sample(c(TRUE, FALSE), n_pop,
                           replace = TRUE, prob = c(0.4, 0.6))
)

# One draw per agent → one realised trajectory each
pop_traj <- sample_trajectories(
  fit, stan_data, pop,
  n_draws = 1,
  seed    = 2026
)

cat(sprintf("Population: %d agents, %d total days of data\n",
            n_distinct(pop_traj$agent_id), nrow(pop_traj)))
```

### Population summary

```{r plot-pop-summary}
#| fig-height: 5

pop_summary <- pop_traj |>
  group_by(agent_id) |>
  summarise(
    peak_rna          = max(log_rna),
    peak_pfu          = max(log_pfu),
    detectable_days   = n(),
    ever_symptomatic  = any(symptomatic == 1),
    lfd_positive_days = sum(lfd),
    .groups = "drop"
  )

pop_summary |>
  select(peak_rna, peak_pfu, detectable_days, lfd_positive_days) |>
  pivot_longer(everything(), names_to = "measure", values_to = "value") |>
  mutate(measure = recode(measure,
    peak_rna = "Peak RNA (log)",
    peak_pfu = "Peak PFU (log)",
    detectable_days = "Detectable days",
    lfd_positive_days = "LFD-positive days"
  )) |>
  ggplot(aes(value)) +
  geom_histogram(bins = 25, fill = "steelblue", alpha = 0.7) +
  facet_wrap(~measure, scales = "free") +
  labs(x = NULL, y = "Agents",
       title = sprintf("Population-level summary (%d agents)", n_pop)) +
  theme_minimal(base_size = 13)
```


## Function reference

```r
sample_trajectories(
  fit,               # CmdStanMCMC fit object
  stan_data,         # Stan data list (from build_stan_data)
  agents,            # data.frame of agent characteristics
  n_draws    = 100,  # posterior draws per agent
  dt         = 1,    # time step (days)
  t_range    = c(-20, 35),  # evaluation window (days from peak)
  lod_rna    = NULL, # RNA LOD (default: Ct 40, NBA calibration)
  lod_pfu    = NULL, # PFU LOD (default: 2.3 ≈ 10 PFU/mL)
  include_noise = TRUE,  # add measurement noise?
  seed       = NULL  # for reproducibility
)
```

### Agent characteristics

| Column            | Values                                                  | Default (ref)  |
|-------------------|---------------------------------------------------------|----------------|
| `age_group`       | `"0-29"`, `"30-49"`, `"50+"`                            | `"0-29"`       |
| `variant`         | `"prealpha"`, `"alpha"`, `"delta"`, `"omicron"`, `"ba4ba5"`, `"other"` | `"prealpha"` |
| `vaccination`     | `"unvaccinated"`, `"boosted"`, `"unboosted"`, `"vaccinated_unreported"`, `"unvaccinated_unreported"`, `"unreported_primary"` | `"unvaccinated"` |
| `prior_infection` | `TRUE` / `FALSE`                                        | `FALSE`        |

Missing columns default to the reference category. Extra columns (e.g. `label`)
are carried through to the output for easy grouping.

### Output columns

| Column                 | Type    | Description                        |
|------------------------|---------|------------------------------------|
| `agent_id`             | int     | Row index of `agents`              |
| `draw`                 | int     | Posterior draw index               |
| `day`                  | numeric | Days relative to RNA peak          |
| `day_since_detectable` | numeric | Days since first detection (≥ 0)   |
| `log_rna`              | numeric | True log RNA (ln copies/mL)        |
| `log_pfu`              | numeric | True log PFU/mL                    |
| `rna_detectable`       | logical | Is true RNA above LOD?             |
| `pfu_detectable`       | logical | Is true PFU above LOD?             |
| `rna`                  | numeric | Measured log RNA                   |
| `pfu`                  | numeric | Measured log PFU                   |
| `lfd`                  | int/num | LFD result (0/1 or probability)    |
| `symptomatic`          | int/num | Symptomatic (0/1 or daily hazard)  |
